---
title: "ASSIGNMENT 05.1: Query a Database with SQL"
author: "Rashaad Mohammed Mirza"
date: "2024-06-10"
output: html_notebook
---

<br/>

**mirza.ra@northeastern.edu**

**CS 5200 – DBMS Su24 | Dr. Martin Schedlbauer**

**Northeastern University**

```{r}
# Load the DBI package, which provides a database interface definition for communication between R and DBMS
library(DBI)
# Load the RSQLite package
library(RSQLite)

# Connect to the OrdersDB.sqlitedb.db database in the current working directory
dbcon <- dbConnect(RSQLite::SQLite(), dbname = "OrdersDB.sqlitedb.db")

# Verify the connection by listing the tables in the database
tables <- dbListTables(dbcon)
print(tables)

# Disconnect from the database when done
# dbDisconnect(dbcon)
```

### 1. What are the last name, first name, and birth date of all employees, sorted by last name?

```{sql connection=dbcon}
-- the connection parameter within the {sql} chunk tells R Markdown which database connection to use for executing the SQL query

SELECT LastName, FirstName, BirthDate
FROM Employees
ORDER BY LastName;
```

### 2. What are the names of each category and the number of distinct products in that category?

```{sql connection=dbcon}
SELECT c.CategoryName, COUNT(DISTINCT p.ProductID) AS ProductCount
FROM Categories c
JOIN Products p ON c.CategoryID = p.CategoryID
GROUP BY c.CategoryName;
```

### 3. List the number of orders by customer ID and name, ordered from most to least, restricted to those customers who have placed at least one order.

```{sql connection=dbcon}
SELECT c.CustomerID, c.CustomerName, COUNT(o.OrderID) AS OrderCount
FROM Customers c
JOIN Orders o ON c.CustomerID = o.CustomerID
GROUP BY c.CustomerID, c.CustomerName
HAVING COUNT(o.OrderID) > 0
ORDER BY OrderCount DESC
LIMIT 10;
```

### 4. What is the last name and first name of the youngest employee?

```{sql connection=dbcon}
SELECT LastName, FirstName, BirthDate
FROM Employees
ORDER BY BirthDate DESC
LIMIT 1;
```

### 5. What are the total number of unique purchases for each country, ordered alphabetically by country, i.e., the number of unique purchases of customers in each country.

```{sql connection=dbcon}
SELECT c.Country, COUNT(DISTINCT o.OrderID) AS UniquePurchases
FROM Customers c
JOIN Orders o ON c.CustomerID = o.CustomerID
GROUP BY c.Country
ORDER BY c.Country
LIMIT 10;
```

### 6. How many employees never worked with a single customer, i.e., they appear in the Employees table but not in the Orders table?

```{sql connection=dbcon}
SELECT COUNT(e.EmployeeID) AS EmployeesWithoutOrders
FROM Employees e
LEFT JOIN Orders o ON e.EmployeeID = o.EmployeeID
WHERE o.OrderID IS NULL;
```

```{sql connection=dbcon}
SELECT (CONCAT(e.FirstName, ' ', e.LastName)) AS EmployeesWithoutOrders, e.EmployeeID
FROM Employees e
LEFT JOIN Orders o ON e.EmployeeID = o.EmployeeID
WHERE o.OrderID IS NULL;
```

### 7. How many products contain either "Caviar" or "Kaviar" in the name of the product?

```{sql connection=dbcon}
SELECT COUNT(*) AS ProductCount
FROM Products
WHERE LOWER(ProductName) LIKE '%caviar%' OR LOWER(ProductName) LIKE '%kaviar%';
```

```{sql connection=dbcon}
SELECT *
FROM Products
WHERE LOWER(ProductName) LIKE '%caviar%' OR LOWER(ProductName) LIKE '%kaviar%';
```

### 8. What is the total number of distinct countries to which at least one order was shipped?

```{sql connection=dbcon}
SELECT COUNT(DISTINCT c.Country) AS ShippedCountries
FROM Orders o
JOIN Customers c ON o.CustomerID = c.CustomerID;
```

```{sql connection=dbcon}
SELECT DISTINCT c.Country AS ShippedCountries
FROM Orders o
JOIN Customers c ON o.CustomerID = c.CustomerID
ORDER BY c.Country
LIMIT 10;
```

### 9. What is the total for each order, i.e., the amount that the customers is required to pay for the products at the ordered quantity in the order. List the order ID, the order date, the name of the customer, and the total cost.

```{sql connection=dbcon}
SELECT
  o.OrderID,
  o.OrderDate,
  c.CustomerName,
  SUM(od.Quantity * p.Price) AS TotalCost
FROM Orders o
JOIN OrderDetails od ON o.OrderID = od.OrderID
JOIN Products p ON od.ProductID = p.ProductID
JOIN Customers c ON o.CustomerID = c.CustomerID
GROUP BY o.OrderID, o.OrderDate, c.CustomerName
ORDER BY o.OrderID
LIMIT 10;
```

### 10. Which supplier sells the most products?

```{sql connection=dbcon}
WITH ProductCounts AS (
  SELECT s.SupplierID, s.SupplierName, COUNT(p.ProductID) AS ProductCount
  FROM Suppliers s
  JOIN Products p ON s.SupplierID = p.SupplierID
  GROUP BY s.SupplierID, s.SupplierName
  ORDER BY s.SupplierID
)

SELECT SupplierID, SupplierName, ProductCount
FROM ProductCounts
WHERE ProductCount = (SELECT MAX(ProductCount) FROM ProductCounts);
```

### 11. Which product was ordered most often by different customers?

```{sql connection=dbcon}
WITH CustomerCounts AS (
  SELECT
    od.ProductID,
    p.ProductName,
    COUNT(o.CustomerID) AS CustomerCount
  FROM OrderDetails od
  JOIN Products p ON od.ProductID = p.ProductID
  JOIN Orders o ON od.OrderID = o.OrderID
  GROUP BY od.ProductID, p.ProductName
  ORDER BY CustomerCount DESC
)

SELECT ProductID, ProductName, CustomerCount
FROM CustomerCounts
WHERE CustomerCount = (SELECT MAX(CustomerCount) FROM CustomerCounts)
ORDER BY ProductID;
```

### 12. Which product has the most quantity sold?

```{sql connection=dbcon}
WITH Product AS (
  SELECT od.ProductID, p.ProductName, SUM(od.Quantity) AS TotalSold
  FROM OrderDetails od
  JOIN Products p ON od.ProductID = p.ProductID
  GROUP BY od.ProductID
  ORDER BY TotalSold DESC
)

SELECT ProductID, ProductName, TotalSold
FROM Product
WHERE TotalSold = (SELECT MAX(TotalSold) FROM Product);
```

### 13. What is the total amount spent by all customers who do live in Germany?

```{sql connection=dbcon}
SELECT SUM(od.Quantity*p.Price) AS TotalAmountSpent
FROM Customers c
JOIN Orders o ON c.CustomerID = o.CustomerID
JOIN OrderDetails od ON o.OrderID = od.OrderID
JOIN Products p ON od.ProductID = p.ProductID
WHERE c.Country = 'Germany';
```

### 14. What is the difference in spending between the country to which the most was sold versus the country to which the least was sold?

```{sql connection=dbcon}
WITH CountrySpending AS (
  SELECT c.Country, SUM(od.Quantity * p.Price) AS TotalSpending
  FROM Customers c
  JOIN Orders o ON c.CustomerID = o.CustomerID
  JOIN OrderDetails od ON o.OrderID = od.OrderID
  JOIN Products p ON od.ProductID = p.ProductID
  GROUP BY c.Country
)

SELECT ((SELECT MAX(TotalSpending) FROM CountrySpending) - (SELECT MIN(TotalSpending) FROM CountrySpending)) AS SpendingDifference;

```

### 15. Which country has the most customers, i.e., which country has the most customers living in it?

```{sql connection=dbcon}
WITH RegionSales AS (
  SELECT c.Country, COUNT(c.CustomerID) AS CustomerCount
  FROM Customers c
  GROUP BY c.Country
  ORDER BY CustomerCount DESC
)

SELECT Country, CustomerCount
FROM RegionSales
WHERE CustomerCount = (SELECT MAX(CustomerCount) FROM RegionSales);
```

### 16. Which employee sold the most in terms of "dollar amount" sold?

```{sql connection=dbcon}
WITH EmpSales AS (
  SELECT
    e.EmployeeID,
    (CONCAT(e.FirstName, ' ', e.LastName)) AS EmployeeName,
    SUM(od.Quantity*p.Price) AS Sales
  FROM Employees e
  JOIN Orders o ON e.EmployeeID = o.EmployeeID
  JOIN OrderDetails od ON o.OrderID = od.OrderID
  JOIN Products p ON od.ProductID = p.ProductID
  GROUP BY e.EmployeeID
  ORDER BY Sales DESC
)

SELECT EmployeeID, EmployeeName, Sales
FROM EmpSales
WHERE Sales = (SELECT MAX(Sales) FROM EmpSales);
```

### 17. Which customers (name and ID) have never bought anything?

```{sql connection=dbcon}
SELECT c.CustomerName, c.CustomerID
FROM Customers c
LEFT JOIN Orders o ON c.CustomerID = o.CustomerID
WHERE o.OrderID IS NULL
LIMIT 10;
```

### 18. What is the average cost of the products in each category? List the category name and the average cost.

```{sql connection=dbcon}
SELECT c.CategoryName, AVG(p.Price) AS AverageCost
FROM Categories c
JOIN Products p ON c.CategoryID = p.CategoryID
GROUP BY c.CategoryName;
```

### 19. What is the most recent order? List the order ID and the name, city, and country of the customer who placed it.

```{sql connection=dbcon}
SELECT o.OrderID, c.CustomerName, c.City, c.Country, o.OrderDate
FROM Orders o
JOIN Customers c ON o.CustomerID = c.CustomerID
ORDER BY o.OrderDate DESC
LIMIT 1;
```

### 20. From how many different (unique) countries do the products come from?

```{sql connection=dbcon}
SELECT COUNT(DISTINCT s.Country) AS SupplierCountries
FROM Suppliers s;
```

```{sql connection=dbcon}
SELECT DISTINCT(s.Country) AS SupplierCountries
FROM Suppliers s
ORDER BY SupplierCountries
LIMIT 10;
```

```{r}
# Disconnect from the database when done
dbDisconnect(dbcon)
```

Note: As per the assignment instructions, many SQL chunks include the LIMIT 10 clause to restrict the result to 10 rows if the query results in more than 10 rows.
